Descriptive Complexity for Counting Complexity Classes

### 2017-09-27 LMCS/proof-prop-rest.tex
[journal] Simplificacion en la demo de Proposicion 5.3 (SigmaQSOSigma0 y #Sigma1 son imcomparables)

We give this proof in three parts.
First, we show that $\E{1} \not\subseteq \QE{0}$. By contradiction, let $\R$ be the empty signature and suppose that there is a $\QE{0}$ formula $\alpha$ over $\R$ such that defines the following function: for every finite $\R$-structure with $n$ elements, $\alpha(\enc(\A)) = n - 1$. We need the following claim.
\begin{clm}
	Let $\alpha = \sa{\bar{x}}\varphi(\bar{x})$	where $\varphi$ is quantifier free. Then the function defined by $\alpha$ is either null, greater or equal to $n$, or is in $\Omega(n^2)$.
\end{clm}
\proof
	Suppose that the function defined by $\alpha$ is not $0$ and that $\varphi$ is in DNF, namely, $\alpha = \sa{\bar{x}} \varphi_1(\bar{x}) \vee \cdots \vee \varphi_n(\bar{x})$. Furthermore, suppose $\bar{x} = (x_1,\ldots,x_{\length{\bar{x}}})$. Since $\alpha$ is not null, then some $\varphi_i$ must be satisfiable and suppose that $\varphi_i(\bar{x}) = \bigwedge\psi(\bar{x})$ where each $\psi(\bar{x})$ is a literal (i.e. an atomic formula or its negation).
	We will prove by induction on $\length{\bar{x}}$ that the function defined by $\sa{\bar{x}} \bigwedge\psi(\bar{x})$ is either greater or equal to $n$, or in $\Omega(n^2)$.
	For $\length{\bar{x}}= 1$ and $\alpha = \sa{x}\bigwedge\psi(x)$, if any $\psi(x) = (x = x)$ or $\neg(x < x)$, then we can eliminate it and we obtain the same function. If any $\psi = (x < x)$ or $\neg(x=x)$, then the function becomes null. The only possible $\alpha$ left is $\alpha = \sa{x}\top$ which is equal to the function $n$. This covers all possible cases for $\length{\bar{x}} = 1$. 
	Now suppose that it holds for $\length{\bar{x}} = k$ and suppose $\alpha = \sa{\bar{x}}\bigwedge\psi(\bar{x})$ for $\length{\bar{x}} = k+1$. If any $\psi(\bar{x}) = (x_i = x_j)$ where $i \neq j$, then $\alpha$ describes the same function as $\alpha$ where $x_j$ has been replaced by $x_i$. In this formula the tuple of first-order variables has now $k$ elements so, by induction, the function satisfies our conclusions. Furthermore, if any $\psi(\bar{x}) = (x_i = x_j)$ or $\neg (x_i < x_j)$ where $i = j$, then we can eliminate $\psi(\bar{x})$ and obtain the same function, and if any $\psi(\bar{x}) = \neg(x_i = x_j)$ or $(x_i < x_j)$ where $i = j$, then the function becomes null. 
	The remaining formulas in $\bigwedge\psi(\bar{x})$ are either $\neg(x_i = x_j)$, $(x_i<x_j)$ or $\neg(x_i<x_j)$ with $i \neq j$. If the formula violates transitivity in $<$ (e.g. $x < y \wedge y < z \wedge z < x$), then $\alpha$ defines the null function. Therefore, there must be some order over $\bar{x}$ that satisfies $\bigwedge\psi(\bar{x})$ and this order can be chosen in  $\binom{n}{\length{\bar{x}}}$ different ways over the domain of $\A$. This implies that $\sem{\alpha} \in \Omega(n^{\length{\bar{x}}}) \subseteq \Omega(n^2)$ if $\length{\bar{x}} > 1$ and concludes the proof of the claim.
\qed
Coming back on proving that $\alpha$ does not define the $(n-1)$-function, suppose that $\alpha$ is in SNF, namely, $\alpha = \sum_{i = 1}^k \alpha_i$ for some fix $k$. Since $\alpha$ is not null, consider some $\alpha_i$ that describes a non-null function. Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$ where $\varphi$ is quantifier-free. Note that if $\length{\bar{X}} > 0$, then the function $\sem{\alpha}$ is in $\Omega(2^n)$, as it was proven in~\cite{SalujaST95}. Therefore, we have that $\alpha_i = \sa{\bar{x}}\varphi(\bar{x})$ and, as we proved in the above claim, it describes either some function greater or equal to $n$, or in $\Omega(n^2)$, which leads to a contradiction. 
To conclude, note that the formula $\sa{x} \ex{y} (x < y)$ is in $\E{1}$ and describes the function $n-1$.

>>>>

We give this proof in three parts.
First, we show that $\E{1} \not\subseteq \QE{0}$. Towards a contradiction, let $\R = \{ < \}$ and suppose that there is a $\QE{0}$ formula $\alpha$ over $\R$ which is equivalent to the $\E{1}$ formula $\sa{x} \ex{y} (x < y)$. This is, for every finite $\R$-structure $\A$ with $n$ elements, $\sem{\alpha}(\A) = n - 1$. We will use the following claim.
\begin{clm}
	Let $\alpha = \sa{\bar{x}}\varphi(\bar{x})$	where $\varphi$ is quantifier free. Then $\alpha$ defines a function that is either null, greater or equal to $n$, or is in $\Omega(n^2)$.
\end{clm}
\proof
	Note that each atomic sub-formula in $\varphi(\bar{x})$ has either the form $(x = y)$, $(x < y)$, $\top$ or a negation thereof, where $x,y\in\bar{x}$. Suppose $\sem{\alpha}$ is not null and consider some $\R$-structure $\A$ such that $\sem{\alpha}(\A) > 0$. Let $\bar{a}$ be an assignment to $\bar{x}$ such that $\A\models\varphi(\bar{a})$. It can be seen that each assignment $\bar{a}'$ that has the same order over its variables\footnote{For example, if $\bar{a} = (a_1,a_2,a_3,a_4)$, such order may be $a_1 > a_3 = a_4 > a_2$.} as $\bar{a}$ also satisfies $\A\models\varphi(\bar{a}')$. If this order has $k$ partitions then $\binom{n}{k}$ assignments for $\bar{x}$ satisfy this order, and therefore, $\sem{\alpha} \geq \binom{n}{k}$. If $k = 1$, then $\binom{n}{k} = n$, and if $k \geq 2$, then $\binom{n}{k} \in \Omega(n^2)$, which proves the claim.
\qed
To prove that $\alpha$ does not define the $(n-1)$-function, suppose that $\alpha$ is in SNF, namely, $\alpha = \sum_{i = 1}^k \alpha_i$ for some fix $k$. Since $\alpha$ is not null, consider some $\alpha_i$ that describes a non-null function. Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$ where $\varphi$ is quantifier-free. Note that if $\length{\bar{X}} > 0$, then the function $\sem{\alpha}$ is in $\Omega(2^n)$, as it was proven in~\cite{SalujaST95}. Therefore, we have that $\alpha_i = \sa{\bar{x}}\varphi(\bar{x})$ and, as we proved in the claim above, it describes either some function greater or equal to $n$, or in $\Omega(n^2)$, which leads to a contradiction.
